Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-01 main.tex
Update on Overleaf.
close reading of evaluation up to SPGEMM

% Configuration for minted package
\setminted{
    style=colorful, % Set code style to colorful
    fontsize=\scriptsize,
    linenos, % Enable line numbers
    breaklines, % Enable line breaks
    breakanywhere, % Allow line breaks anywhere
    autogobble,
}
\setmintedinline{fontsize=\small}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

>>>>

% Configuration for minted package
\setminted{
    style=colorful, % Set code style to colorful
    fontsize=\tiny,
    linenos=false, % Enable line numbers
    breaklines, % Enable line breaks
    breakanywhere, % Allow line breaks anywhere
    autogobble,
}
\setmintedinline{fontsize=\small}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

### 2024-04-01 sections/eval.tex
add reproducibility paragraph
close reading of evaluation up to SPGEMM

\section{Evaluation}

@@

\subsection{Data-Driven Performance Engineering}
\subsubsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
%\changwan{
%Explain what is SpGEMM. why SpGEMM hard? (not sure)
%three approaches. Asym time complexity?
%}

Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics, focusing on efficiently computing the product of two sparse matrices to produce an output sparse matrix.

The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
Further explanation is available in Section 2.2 of~\cite{zhang2021gamma}.

Figures~\ref{fig:spgemm_small} and~\ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO. Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals. Additionally, TACO only supports outer product when the output matrix format is dense. It does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices. So, we were unable to obtain numbers for inner-product and TACO's outer-product for large matrices. In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.

As depicted in Figure~\ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product. For large sparse matrices, as shown in Figure~\ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05. It's worth mentioning that TACO's Gustavson's algorithm is hard-wired whereas Finch's data structure-driving programming enables to systematically generate SpGEMM using explicit temporary, bytemap format, and matrix transposition.

%\help{The main points of this section are:}

%\help{1. summarize the difference between inner, gustavson, and outer, cite gamma but this also needs to stand on it's own}

%\help{2. Explain why bytemap temporary makes sense for gustavson, but not for outer or inner}

%\help{3. Point out that TACO does not even support outer product because it does not support scatter into 2D workspace}

%\help{4. Take a look at the figure and point out that Finch is competitive with TACO on inner and gustavson, and that for outer product we see what we would expect from TACO: i.e. Finch Outer with dense performs similarly to TACO with dense, and Finch Outer with sparse performs much better as the matrix size increases.}

%\help{5. point out that for the large matrices finch is competitive with TACO for gustavson}.

%\help{6. explain the features of "datastructure-driven programming" that are used here in this problem. i.e. explicit temporary, bytemap format, transpositions}.
\begin{figure}
    \label{fig:spgemm}
	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
    \caption{A comparison of several matrix multiplication algorithms between
    Finch and Taco. On left, we use smaller matrices, ordered from small to big
    dimension. Note that inner products necessarily requires $O(n^2)$ work and
    taco's outer products format is dense. Finch can use a sparse outer products
    format and thus has an asymptotic advantage that becomes evident as the
    output dimensions grow. On right, we use only gustavson's algorithm and compare on larger matrices.}
\end{figure}

Examples that demonstrate performance engineering in a datastructure-driven model

\subsubsection{SpMV}
The sparse matrix-vector multiplication kernel (SpMV) is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite]. SpMV is a bandwidth bound kernel, and thus historically has been subject to many modifications to improve throughput and reduce the associated control flow [cite]. The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.

@@

\subsection{Programming over flexible data}

\subsubsection{Image Morphology}

@@

\subsubsection{Graph Analytics}
%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e. choose), and that Finch can do early break, and also the different loop orders and multiple outputs. We may need to explain a little bit about what push pull is. For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}

>>>>

\section{Case Studies}

@@

All experiments were run on a single core of a 12-core 2-socket Intel Xeon E5-2695 v2 running at
2.40GHz with 128GB of memory. Finch is implemented in Julia v1.9, targeting LLVM
through Julia. All timings are the minimum of 10,000 runs or 5s of measurement,
whichever happens first.

\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
\begin{figure}
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    @finch begin
      C .= 0
      for j=_
        for i=_
          for k=_
            C[i, j] += AT[k, i] * B[k, j]
          end
        end
      end
      return C
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    w = Tensor(SparseByteMap(Element(0)))
    @finch begin
      C .= 0
      for j=_
        w .= 0
        for k=_
          for i=_
            w[i] += A[i, k] * B[k, j]
          end
        end
        for i=_
          C[i, j] = w[i]
        end
      end
    end
    \end{minted}
    \end{minipage}%
    \begin{minipage}{0.333\linewidth}
    \begin{minted}{julia}
    w = Tensor(SparseHash(SparseHash(Element(0))))
    @finch begin
      w .= 0
      for k=_
        for j=_
          for i=_
            w[i, j] += A[i, k] * BT[j, k]
          end
        end
      end
      C .= 0
      for j=_, i=_
        C[i, j] = w[i, j]
      end
    end
    \end{minted}
    \end{minipage}
    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
\end{figure}

Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.

There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
%
The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
%
Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
%
Inner products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.

Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
%
Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
%
Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
%
In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
%
Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
%
It's also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.

As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures. For large sparse matrices, Finch outperforms TACO with an average speedup of 1.05.

\begin{figure}
	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
    \caption{A comparison of several matrix multiplication algorithms between
    Finch and Taco. On left, we use smaller matrices, ordered from small to big
    dimension. Note that inner products necessarily requires $O(n^2)$ work and
    TACO's outer products format is dense. Finch can use a sparse outer products
    format and thus has an asymptotic advantage that becomes evident as the
    output dimensions grow. On right, we use only gustavson's algorithm and
    compare on larger matrices.}
    \label{fig:spgemm}
\end{figure}

\subsection{SpMV}
The sparse matrix-vector multiplication kernel (SpMV) is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite]. SpMV is a bandwidth bound kernel, and thus historically has been subject to many modifications to improve throughput and reduce the associated control flow [cite]. The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.

@@

\subsection{Image Morphology}

@@

\subsection{Graph Analytics}
%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e. choose), and that Finch can do early break, and also the different loop orders and multiple outputs. We may need to explain a little bit about what push pull is. For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}
